---
title: "DATA1001 Project"
author: "SIDs: 560654992"
format: 
  html:
    embed-resources: true
    toc: true
    code-fold: true
    theme: darkly
    
---

# Executive Summary (Max 50 words)

Older demographics, between 40 - 45 tend to report higher standard alcohol intake and is more notable for males and especially non-binaries. Furthermore, older demographics generally report more past dates suggesting more accumulated relationship experience. Furthermore the distribution of data for this is heavily influenced by gender as reported females contribute to the strong variation observed in the data. 

# Evidence

## EDA (Max 200 words)

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
datasurvey <- read_csv("/Users/sebastian/Downloads/Data Sci/data1001_survey_data_2026S1-1.csv")
```


Discuss the data and consider

1.  Source
2.  Structure
3.  Limitations

# Source

Variables focused on:

Age – Qualitative (ordinal)

Gender – Qualitative (nominal)

Standard Drinks – Quantitative (continuous)


# Limitations


# Assumptions
We assumed that most responses were honest and reasonable in answering the questions,  During data cleaning, we removed entries with missing age (eg, N/A) information and excluded records reporting more than 20 standard drinks, as such values exceed safe consumption limits and were signs of untruthful responses. Age values were then grouped into five‑year intervals ranging from 15 to 80 years, which represented the most appropriate bounds based on the distribution of the dataset.

## 1) How does age relate to standard alcohol intake, and does this relationship differ by gender? (Max 350 words for both RQ1 + RQ2 combined)

Type your research question evidence here.

```{r fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(readr)

datasurvey <- read_csv("/Users/sebastian/Downloads/Data Sci/data1001_survey_data_2026S1-1.csv", na = c("", "NA", "N/A", "n/a", "na", "NaN", "null", "NULL"))

clean_data <- datasurvey %>%
  mutate(across(where(is.character), ~str_trim(.x))) %>%
  mutate(across(where(is.character), ~na_if(.x, "n/a"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "N/A"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "NA"))) %>%
  mutate(age = suppressWarnings(as.numeric(age)),
         standard_drinks = suppressWarnings(as.numeric(standard_drinks))) %>%
  filter(is.finite(age),
         is.finite(standard_drinks),
         !is.na(gender),
         gender != "") 

plot_data <- clean_data %>%
  filter(gender %in% c("Female", "Male", "Non-binary / third gender")) %>%
  group_by(gender, age) %>%
  summarise(median_drinks = median(standard_drinks, na.rm = TRUE), .groups = "drop")

alcoholxage <- ggplot(plot_data, aes(x = age, y = median_drinks, color = gender)) +
  geom_jitter(width = 1, alpha = 0.35) +
  geom_smooth(method = "lm", linewidth = 1, se = FALSE) +
  labs(title = "Median Standard Drinks by Age Group and Gender",
       x = "Age",
       y = "Median Standard Drinks") +
  scale_x_continuous(
       limits = c(15, 80),
       breaks = seq(15, 80, by = 5)) +
  scale_y_continuous(
       limits = c(0, 20)
  )
  theme_minimal()

ggplotly(alcoholxage)
```

## Research Question 2 (Linear Model)
# “To what extent do age and gender predict the usage of drugs reported by students?”
```{r}
```{r fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(readr)

datasurvey <- read_csv("/Users/sebastian/Downloads/Data Sci/data1001_survey_data_2026S1-1.csv", na = c("", "NA", "N/A", "n/a", "na", "NaN", "null", "NULL"))

breaks <- seq(15, 80, by = 5)
labs <- paste0(breaks[-length(breaks)], "–", breaks[-1])

clean_data <- datasurvey %>%
  mutate(across(where(is.character), ~str_trim(.x))) %>%
  mutate(across(where(is.character), ~na_if(.x, "n/a"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "N/A"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "NA"))) %>%
  mutate(age = suppressWarnings(as.numeric(age)),
         standard_drinks = suppressWarnings(as.numeric(standard_drinks))) %>%
  filter(is.finite(age),
         is.finite(standard_drinks),
         !is.na(gender),
         gender != "") %>%
  mutate(age_group = cut(age,
                     breaks = breaks,
                     right = FALSE,
                     include.lowest = TRUE,
                     labels = labs)) %>%
  filter(!is.na(age_group))

plot_data <- clean_data %>%
  filter(gender %in% c("Female", "Male", "Non-binary / third gender")) %>%
  group_by(gender, age_group) %>%
  summarise(median_drinks = median(standard_drinks, na.rm = TRUE), .groups = "drop")

alcoholxage <- ggplot(plot_data, aes(x = age_group, y = median_drinks, fill = gender)) +
  geom_col(show.legend = FALSE, width = 0.8) +
  facet_wrap(~ gender, nrow = 1) +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  labs(title = "Median Standard Drinks by Age Group and Gender",
       x = "Age Group",
       y = "Median Standard Drinks") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.title.x = element_text(margin = margin(t = 12)),
        panel.spacing = unit(1.2, "lines")) +
  guides(x = guide_axis(n.dodge = 2)) 

ggplotly(alcoholxage)
```
```

Type your research question evidence here. Include testing assumptions of the linear model (residual plot).


## Articles (Max 50 words)

Include any relevant literature to your analysis.

# Acknowledgements

The Acknowledgment section includes a list of group meetings (date and time and attendance), the contribution of each group member, and all resources used (e.g. url of stack overflow, url of Ed post, date and details of drop-in session with tutor). 

An AI usage statement including: 
Codex - Chatgpt 5.3: Used to increase spacing in the x-axis and clean N/A values

A short statement explaining how 1 of the Shared Values and 1 of the Ethical Principles has been adhered to.