---
title: "DATA1001 Project"
author: "SIDs: 560654992"
format: 
  html:
    embed-resources: true
    toc: true
    code-fold: true
    theme: darkly
    
---

# Recommendation/Insight

Older demographics, between 40 - 45 tend to report higher standard alcohol intake and is more notable for males and especially non-binaries. Furthermore, older demographics generally report more past dates suggesting more accumulated relationship experience. Furthermore the distribution of data for this is heavily influenced by gender as reported females contribute to the strong variation observed in the data.

# Evidence

## EDA (Max 200 words)

Discuss the data and consider

1.  Source
2.  Structure
3.  Limitations

# Source

Variables focused on was sourced from a survey containing 26 testable variables:

**Age** – Qualitative (ordinal)

**Gender** – Qualitative (nominal)

**Drug Usage** - Qualitative (nominal)

**Standard Drinks** – Quantitative (continuous)


# Limitations

Not every person in DATA1001 consented to participate in the study, so the dataset represents a sample rather than the entire population. The question about drug use was randomly assigned to participants, either as “Have you ever used recreational drugs?” or “Have you ever gotten high off illicit drugs?” This introduces a potential confounding variable, because a participant who answers “no” to having gotten high off illicit drugs might still have used recreational drugs at some point, leading to misclassification in the responses.

# Assumptions
We assumed that most responses were honest and reasonable in answering the questions,  During data cleaning, we removed entries with missing age (eg, N/A) information and excluded records reporting more than 20 standard drinks, as such values exceed safe consumption limits and were signs of untruthful responses. Furthermore age values were left from 15 to 80 years, which represented the most appropriate bounds based on the distribution of the dataset.

## 1) Is regular alcohol consumption associated with an increased likelihood of trying recreational drugs? (Max 350 words for both RQ1 + RQ2 combined)
```{r fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(readr)
datasurvey <- read_csv("/Users/sebastian/Downloads/data1001_survey_data_2026S1-1.csv")

clean_data <- datasurvey %>%
    
mutate(
    standard_drinks = suppressWarnings(as.numeric(standard_drinks)),
    drug_use_ans = na_if(drug_use_ans, "Prefer not to say")) %>%
  filter(
    !is.na(drug_use_ans),
    !is.na(standard_drinks))

drugxalcohol <- ggplot(clean_data, aes(y = drug_use_ans, x = standard_drinks, fill = drug_use_ans)) +
  geom_boxplot(alpha = 0.7, width = 0.3) +
  scale_fill_manual(values = c("Yes" = "red", "No" = "green")) +
  scale_x_continuous(limits = c(0, 20)) +
  labs(title = "Median Standard Drinks by Age Group and Gender",
       x = "Median Standard Drinks",
       y = "Drug Use") +
  theme_minimal() + theme(
  plot.title = element_text(size = 20, face = "bold"),
  legend.title = element_text(size = 14),
  legend.text = element_text(size = 14),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 14))

drugxalcohol
```



Type your research question evidence here.

## Research Question 2 (Linear Model)
# How does age relate to standard alcohol intake, and does this relationship differ by gender?

```{r fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(readr)

datasurvey <- read_csv("/Users/sebastian/Downloads/Data Sci/data1001_survey_data_2026S1-1.csv", na = c("", "NA", "N/A", "n/a", "na", "NaN", "null", "NULL"))

breaks <- seq(15, 80, by = 5)
labs <- paste0(breaks[-length(breaks)], "–", breaks[-1])

clean_data <- datasurvey %>%
  mutate(across(where(is.character), ~str_trim(.x))) %>%
  mutate(across(where(is.character), ~na_if(.x, "n/a"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "N/A"))) %>%
  mutate(across(where(is.character), ~na_if(.x, "NA"))) %>%
  mutate(age = suppressWarnings(as.numeric(age)),
         standard_drinks = suppressWarnings(as.numeric(standard_drinks))) %>%
  filter(is.finite(age),
         is.finite(standard_drinks),
         !is.na(gender),
         gender != "") 

plot_data <- clean_data %>%
  filter(gender %in% c("Female", "Male", "Non-binary / third gender")) %>%
  group_by(gender, age) %>%
  summarise(median_drinks = median(standard_drinks, na.rm = TRUE), .groups = "drop")

alcoholxage <- ggplot(plot_data, aes(x = age, y = median_drinks, color = gender)) +
  geom_jitter(width = 1, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(limits = c(15, 80), breaks = seq(15, 80, by = 5)) +
  scale_y_continuous(limits = c(0, 20)) +
  labs(title = "Median Standard Drinks by Age Group and Gender",
       x = "Age",
       y = "Median Standard Drinks") +
  theme_minimal() 

ggplotly(alcoholxage)
```
Type your research question evidence here. Include testing assumptions of the linear model (residual plot).

## Articles (Max 50 words)

Include any relevant literature to your analysis.

# Acknowledgements

The Acknowledgment section includes a list of group meetings (date and time and attendance), the contribution of each group member, and all resources used (e.g. url of stack overflow, url of Ed post, date and details of drop-in session with tutor). 

An AI usage statement including: 
Codex - Chatgpt 5.3: Used to increase spacing in the x-axis and clean N/A values for Research Question 1

A short statement explaining how 1 of the Shared Values and 1 of the Ethical Principles has been adhered to.